Particle Smoothing Algorithms for Variable Rate Models

### 2012-10-11 response_to_reviewers.tex

\usepackage{amsmath}
\usepackage{IEEEtrantools}
\usepackage{color}

\newenvironment{meta}[0]{\color{red} \em}{}

\begin{document}

{\meta Thankyou for the comments, all of which have been addressed. Responses are inline with the individual points below. }

@@

\subsubsection*{Comments}
This is an interesting, well written paper  that suggests new sampling strategies for smoothing  in variable rate models. The proposed method  rejuvenates the variable rate filtering particles that usually have poor smoothing performance due to the loss of particle diversity particularly when state discontinuities occur.

The paper relays on the key idea, using the previously generated filtering particles as a proposal  to the target conditional distributions, which is inherited from the previous work of the author [Godsill et al. 2004] . The proposed algorithms apply a backward pass by sampling the  target conditional distribution $P( \theta^{-}_{n} |  . )$  for $n=N…1$.   Two methodologies are proposed for sampling the target conditional distribution of the  variable rate structured "conditionally linear Gaussian models"  and "conditionally deterministic models". The former samples from a discrete set of re-weighted filtering particles according to well known Rao-Blackwellization approach and the other uses an  MCMC sampling scheme where the proposals are constructed by a  modified  filtering particle set.

Both algorithms are sophisticated techniques for rejuvenating the particles and can be  interpreted as a way of reconstructing the particle history by using the filtering particles sampled at each filtering step. Reference on the convergence property of the algorithms and a  discussion to understand the practical aspects of the the algorithm (Ex. whether is it  possible to increase the online filtering performance by using retrospective smoothing steps, the ability of the algorithm to  find the untracked modes of the posterior - this can also be suported with simulation results )  would be useful  for the reader in order to understand the limits of the algorithms.

{\meta Added a brief note on convergence, which is inherited from the basic results for the forward-filtering-backward-sampling algorithm. Added an explicit example in the simulations of the smoother representing multi-modality where some particle contain a jump and others do not. Fixed interval smoothing will not help online filtering. Fixed-lag smoothing would, and this is covered elsewhere, especially in [Whitely et al. 2011]. }

@@

In the simulations section, smoothing performance of the algorithms are compared to filtering and  the filtering- smoothing results.  In both tests improvement is achieved in means of RMSE. I see that in the  simulations filter and filter- smoother is able to represent the solution to the data. However it would be much more interesting and convincing to see an illustration of a scenario  where the filter is not able to represent the solution due to discarded particles in the resampling steps  (Ex.  poor estimate for the changepoint location or  parameters,  misses a jump or unable to track the maneuver due to multi-modality)  but  the smoother is able track the solution.

{\meta Because the smoother uses filter particles as a proposal, if the filter fails catastrophically (e.g. track loss) then the smoother will also fail. Added an example where the multi-modality of the particle representation is lost due to resampling in the filter-smoother, but not the filter. }

@@

Recommendation: RQ - Review Again After Major Changes

\subsubsection*{Comments}
The authors propose a smoothing PF for variable rate models. The proposed method is essentially an adaptation of existing methods for fixed-rate models to the variable rate model. I feel there is enough substance in the adaptation process to merit publication provided some revisions are performed. My main concerns are with the performance analysis:

- It is never clarified but I'm assuming that the smoothing results are for fixed-interval smoothing, i.e., estimates at time $t_n, n=1,...,N$ use measurements from times $t_1,...,t_N$. This should be clearly stated.

{\meta This should be clear from the introduction and the maths, but it is now explicitly stated as well. }

@@

{\meta Ok. Now 100.}

- In Figs 7 and 10 the effective sample size would be a better measure of performance than the number of unique particles.

- The principal improvement offered by the more sophisticated particle smoothers compared to the filter-smoother is increased particle diversity. This should improve estimation performance but the improvements in consistency may be even greater. This is alluded to in the paper and partially covered by Figs 6 and 9 which show plots of sample trajectories. However, a more rigorous consistency analysis should be performed. For example, a simple approach would be to construct a Gaussian approximation to the posterior, establish a confidence ellipsoid and then count the number of times the true value falls within this ellipsoid.

{\meta These two comments have been addressed by replacing the ``number of particles'' comparison with normalised estimation error squared (NEES) statistics. When the number of particles is very low, as is the case for the filter-smoother, covariance estimates become degenerate and the NEES is infinite. Therefore, a transformed version of the NEES is used. These show a better efficiency of the smoother relative to the filter-smoother. This is based on very similar principles to the test suggested (i.e. comparing Mahalanobis distances).}

@@

- Comparing the proposed method and the filter-smoother for equal sample sizes doesn't really make sense given that the former has $O(N^2)$ expense while the latter has $O(N)$ expense. The filter-smoother should be implemented with a much larger sample size such that both algorithms have approximately equal expense. This approach, which is taken in (Fearnhead et al, 2010), seems much fairer.

{\meta %I disagree with this. Although it might be a fairer test for a given problem, the result can be changed by modifying the scenario. For example, the relative performance of the smoother will be improved by having a longer data set. The objective of the simulations section is to demonstrate that the smoother operates as expected --- producing rejuvenated particle sets. The better algorithm for any application will depend on the model, the length of the data, and the desired output (e.g. is a distributional estimate needed, or is a single sample enough?).

%In contrast to us, [Fearnhead et al. 2010] are comparing a variety of smoothing algorithms all of which do the same thing, produce a diverse set of marginal smoothing particles. Simple changes like altering the data length are less likely to affect their comparisons, so their methods are appropriate to their purpose, but less so here.

%All this said, increasing the number of particles used by the filter-smoother so as to equalise processing time does very little to improve its performance, so we've done it anyway.

Done.

}

@@

- Results should be shown for a range of sample sizes so we can get an idea of the required sample size for (close to) optimal performance.

{\meta Yes. Added graphs showing the effect of varying the number of smoothing trajectories for the tracking case. The finance results are similar and so omitted for space. }

@@

Some minor comments:
- A Rao-Blackwellised VRPF was also developed in (Morelande and Gordon, 2009). Their formulation was more general than that used here since it permits the model to be partially linear-Gaussian.

{\meta Yes. This should be and now is referenced. Actually, their formulation is not more general, just different. The Rao-Blackwellisation in our paper is of the state variables when the dynamics are linear-Gaussian, while [Morelande and Gordon, 2009] Rao-Blackwellise the linear parts of the motion parameters, but in a conditionally deterministic model. We could extend our work by using their Rao-Blackwellisation scheme on the piecewise-deterministic smoother. }

- Recently, O(N) schemes for particle smoothing which improve upon the filter-smoother have been proposed by (Fearnhead et al, 2010) and (Douc et al, 2011). Perhaps these could be used for VR filtering. They should at least be mentioned.

{\meta Added a reference to [Douc et al., 2011]. This is similar to our Metropolis-Hastings-based method which is already referenced (still awaiting publication). ([Bunch and Godsill, 2012]). The Fearnhead method is for marginal (as opposed to joint) smoothing distributions only, so it is not relevant here. This is now mentioned in section IV. }

@@

- In Fig. 9(a), are these results for the filter-smoother?

{\meta Yes. Clarified.}

- I'm not sure all the appendices are necessary since the results can be found in other, easily obtained sources. At least

{\meta I agree. Removed the first two, giving the derivation and initialisation of the two-filter smoother. The discretisation of the finance model remains, as I can't find a complete and equivalent version in a published work. }

### 2012-10-17 second_submit_doublecol_VR_smoothing.tex
Second submission

The target distribution for a variable rate smoothing algorithm is the posterior distribution over the entire changepoint sequence, $p(\theta|y_{1:N})$. This distribution is approximated by the final step of the VRPF. However, in the same manner as the fixed rate filter-smoother of \cite{Kitagawa1996}, this approximation is likely to lack path-space diversity; because of the necessary resampling step in the filtering algorithm, the particles all share the same set of changepoints before a particular time, with variation only appearing for changepoints closer to $T$. For a better characterisation of the smoothing distribution, it is necessary to rejuvenate the set of particles. This is achieved here with a backward pass through the observations in an extension of the forward-filtering-backward-sampling (FFBS) method described in \cite{Godsill2004}. Since these algorithms are based on the FFBS algorithm, the convergence proofs of \cite{Godsill2004,Douc2011} are believed to apply.

@@

The algorithmic complexity of this variable rate particle smoother for conditionally linear-Gaussian models is $O(\aleph_F \times \aleph_S \times N)$. It is possible to reduce this by using a rejection sampling \cite{Douc2011} or MCMC \cite{Bunch2012} sampling scheme, which avoids the necessity of calculating the sampling weights for every filter particle.

@@

Comparisons of accuracy are most easily conducted by comparing state estimates. These can be calculated by simply taking a mean of the particle values (using the Gaussian means for the linear-Gaussian case). The accuracy is then assessed by comparing the size of estimation errors of the resulting estimators. In addition, the benefit of employing a smoothing algorithm is that it ensures a more diverse particle representation of the posterior state distribution. To demonstrate this effect, a measure of estimator consistency is calculated by which to compare the algorithms, the mean empirical normalised estimation error squared (MENEES). This is defined as,  %the number of unique state samples is plotted over time in the following examples. A
%
\begin{IEEEeqnarray}{rCl}
 \text{MENEES}_k & = & (\hat{x}_k - x_k^*)^T \hat{P}_k (\hat{x}_k - x_k^*) \nonumber \\ 
 \hat{x}_k & = & \sum_i x_{k}^{(i)} \nonumber \\
 \hat{P}_k & = & \sum_i (x_{k}^{(i)} - x_k^*) (x_{k}^{(i)} - x_k^*)^T \nonumber     ,
\end{IEEEeqnarray}

where $x_k^*$ is the true state, and $x_k^{(i)}$ are the individual particle states. This statistic takes a value between $0$ and $1$, where $1$ is scored by a degenerate particle estimate (fewer linearly independent particles than state dimensions). For a non-degenerate estimate, smaller values correspond to smaller estimation errors relative to the estimated covariance. Thus, it incorporates the diversity of the particle set as well the accuracy of the estimate and covariance.

@@

The filter used $\aleph_F = 100$ particles, and the smoother resampled $\aleph_S = 100$ sequences. Bootstrap proposals were used for the filter. An example forward realisation simulated from the model is shown in figure~\ref{fig:example_data}. The sampled trend trajectories produced by the filter-smoother and VRPS are shown in figure~\ref{fig:example_state} (the equivalent graphs for value are not particularly informative as this quantity is observed, and the errors produced by both algorithms are small). Notice that the path diversity is dramatically improved by the VRPS.

@@

Performance was evaluated by testing on 100 scenarios, each with 1000 observations. At each observation time, an estimate of the value and trend is made by taking the average of Gaussian means generated by an RTS smoother for each particle. The resulting mean absolute error over all scenarios is shown in figure~\ref{fig:finance_rmse}, demonstrating a small improvement in accuracy from the smoother. More significantly, MENEES is shown in figure~\ref{fig:finance_menees}, showing a similar consistency between filter and smoother. The filter-smoother produces an MENEES of $1$ at almost all time instants, because there is only one unique particle in the set.% are displayed in table~\ref{tab:finance_state_performance}

@@

\includegraphics[width=0.9\columnwidth]{finance_menees.pdf}
\caption{MENEES over 100 scenarios using filter, filter-smoother, and smoother.}
\label{fig:finance_menees}

@@

As well as providing more accurate and more diverse estimates, the smoother is also capable of preserving multi-modality in its representation of the posterior distribution, which is lost by the filter smoother due to the particle collapse to a single value. This is highlighted in figure~\ref{fig:finance_multimodality}, which shows a segment of the trend estimates from figure~\ref{fig:example_state} containing a small jump. In the filter-smoother results, none of the particles have a jump at this point. In contrast, a significant number of the smoother particles contain a jump around about the right place. This ability to represent multi-modality is a key advantage of good particle smoothing algorithms.

@@

The filter and smoother were each used to generate $50$ particles, with the filter employing resample-move steps with proposals based on an unscented transform \cite{Julier2004} approximation to the optimal proposal density, and the smoother using $M=5$ steps in each Markov chain.

@@

Performance was evaluated by testing on 100 independently generated scenarios, each with 500 observations. At each observation time, an estimate of the position and velocity is made by taking the average of values from the array of particles. The resulting mean absolute error over all scenarios is shown in figure~\ref{fig:finance_rmse}, again demonstrating a small improvement in accuracy from the smoother. The smoother also provides a more accurate estimate of the number of changepoints in each test, with an average error of 0.6 compared to 2.0 from the filter-smoother. MENEES is displayed in figure~\ref{fig:tracking_menees}, showing a similar consistency between filter and smoother. The filter-smoother produces an MENEES of $1$ at almost all time instants, because there is only one unique particle in the set. %in table~\ref{tab:tracking_state_performance}, demonstrating an improvement in accuracy from the smoother.

@@

\includegraphics[width=0.9\columnwidth]{tracking_menees.pdf}
\caption{MENEES over 100 scenarios using filter, filter-smoother, and smoother.}
\label{fig:tracking_menees}

@@

Finally, in figure~\ref{fig:varying_num_pts} the variation in accuracy and consistency of the smoothing particle approximation is demonstrated as the number of smoothing particles varies. As expected, the rate of improvement diminishes and becomes negligible as the number of smoothing particles exceeds the number of filtering particles.

@@

@@

\subfloat[]{\includegraphics[width=0.45\columnwidth]{MENEES_vs_numpts.pdf}}
\caption{Average position error (a) and MENEES (b) over 100 scenarios with varying number of smoother particles and 50 filter particles.}

>>>>

The target distribution for a variable rate smoothing algorithm is the posterior distribution over the entire changepoint sequence, $p(\theta|y_{1:N})$. This distribution is approximated by the final step of the VRPF. However, in the same manner as the fixed rate filter-smoother of \cite{Kitagawa1996}, this approximation is likely to lack path-space diversity; because of the necessary resampling step in the filtering algorithm, the particles all share the same set of changepoints before a particular time, with variation only appearing for changepoints closer to $T$. For a better characterisation of the smoothing distribution, it is necessary to rejuvenate the set of particles. This is achieved here with a backward pass through the observations in an extension of the forward-filtering-backward-sampling (FFBS) method described in \cite{Godsill2004}.% Since these algorithms are based on the FFBS algorithm, the convergence results of \cite{Godsill2004,Douc2011} are believed to apply.

@@

The algorithmic complexity of this variable rate particle smoother for conditionally linear-Gaussian models is $O(\aleph_F \times \aleph_S \times N)$. It is possible to reduce this by using a rejection sampling \cite{Douc2011} or MCMC \cite{Bunch2012} sampling scheme, which avoids the necessity of calculating the sampling weights for every filter particle. Since the algorithm is based on the FFBS algorithm, the convergence results of \cite{Godsill2004,Douc2011} are believed to apply.

@@

Comparisons of accuracy are most easily conducted by comparing state estimates. These can be calculated by simply taking a mean of the particle values (using the Gaussian means for the linear-Gaussian case). The accuracy is then assessed by comparing the size of errors of the resulting estimators. In addition, the benefit of employing a smoothing algorithm is that it ensures a more diverse particle representation of the posterior state distribution. To demonstrate this effect, a measure of efficiency is calculated by which to compare the algorithms. The normalised estimation error squared (NEES) \cite{Bar-Shalom1995,Briers2003} is a statistic derived from the chi-squared test. It is defined by,
%
\begin{IEEEeqnarray}{rCl}
 \text{NEES}_k & = & (\hat{x}_k - x_k^*)^T \hat{P}_k^{-1} (\hat{x}_k - x_k^*) \nonumber \\
 \hat{x}_k & = & \frac{1}{N_S} \sum_i x_{k}^{(i)} \nonumber \\
 \hat{P}_k & = & \frac{1}{N_S} \sum_i (x_{k}^{(i)} - \hat{x}_k) (x_{k}^{(i)} - \hat{x}_k)^T \nonumber     ,
\end{IEEEeqnarray}

where $x_k^*$ is the true state value and $x_k^{(i)}$ are the individual particle states. Calculating NEES for particle approximations which might be degenerate is problematic, because $\hat{P}_k$ will not be invertible (resulting in an infinite value). Here we map the interval $(0, \infty)$ onto $(0, 1)$ using the following transformation, giving a transformed NEES (TNEES) score.
%
\begin{IEEEeqnarray}{rCl}
 \text{TNEES}_k & = &\frac{\text{NEES}_k}{1+\text{NEES}_k}     .
\end{IEEEeqnarray}

Assuming Gaussian distributed estimation errors, it can be shown that NEES should obey a chi-squared distribution with the degrees of freedom parameter equal to the number of state dimensions. Under this assumption, an optimal value of mean TNEES can be found using the transformation above and compared to the average over a large number of Monte Carlo runs. The integral required to calculate this mean is intractable but may be accurately approximated by simulation.

@@

The filter and filter-smoother were run using using $\aleph_F = 1000$ particles, while the smoother used $\aleph_F = 100$ particles and was used to simulate $\aleph_S = 100$ smoothed paths. These values lead to roughly equal running time for the smoother and filter-smoother. Bootstrap proposals were used for the filter. An example forward realisation simulated from the model is shown in Fig.~\ref{fig:example_data}. The sampled trend trajectories produced by the filter-smoother and VRPS are shown in Fig.~\ref{fig:example_state} (the equivalent graphs for value are not particularly informative as this quantity is observed, and the errors produced by both algorithms are small). Notice that the path diversity is dramatically improved by the VRPS.

@@

Performance was evaluated by testing on 100 scenarios, each with 1000 observations. At each observation time, an estimate of the value and trend is made by taking the average of Gaussian means generated by an RTS smoother for each particle. The resulting mean absolute error over all scenarios is shown in Fig.~\ref{fig:finance_rmse}, demonstrating a small improvement in accuracy from the smoother. Mean TNEES is shown in Fig.~\ref{fig:finance_mtnees}, showing a similar efficiency between filter and smoother. The filter-smoother produces a mean TNEES close to $1$ at most time instants, because there is often only one unique particle in the set.% are displayed in table~\ref{tab:finance_state_performance}

@@

\includegraphics[width=0.9\columnwidth]{finance_mtnees.pdf}
\caption{Mean TNEES over 100 scenarios using filter, filter-smoother, and smoother. Optimum value: 0.54}
\label{fig:finance_mtnees}

@@

As well as providing more accurate and more diverse estimates, the smoother is also capable of preserving multi-modality in its representation of the posterior distribution, which is lost by the filter smoother due to the particle collapse to a single value. This is highlighted in Fig.~\ref{fig:finance_multimodality}, which shows the trend estimates from a segment of simulated data containing a small jump. In the filter-smoother results, none of the particles have a jump at this point. In contrast, a significant number of the smoother particles contain a jump around about the right place. This ability to represent multi-modality is a key advantage of good particle smoothing algorithms.

@@

The filter and filter-smoother were run using using $\aleph_F = 500$ particles, while the smoother used $\aleph_F = 50$ particles and was used to simulate $\aleph_S = 50$ smoothed paths. These values lead to roughly equal running time for the smoother and filter-smoother. The filter employed resample-move steps with proposals based on an unscented transform \cite{Julier2004} approximation to the optimal proposal density, and the smoother using $M=5$ steps in each Markov chain.

@@

Performance was evaluated by testing on 100 independently generated scenarios, each with 500 observations. At each observation time, an estimate of the position and velocity is made by taking the average of values from the array of particles. The resulting mean absolute error over all scenarios is shown in Fig.~\ref{fig:tracking_rmse}, again demonstrating a small improvement in accuracy from the smoother. The smoother also provides a more accurate estimate of the number of changepoints in each test, with an average error of 0.6 compared to 2.0 from the filter-smoother. Mean TNEES is displayed in Fig.~\ref{fig:tracking_mtnees}. It shows that while the smoother does not match the efficiency of the filter, it is a significant improvement over the filter-smoother, which has a Mean TNEES of $1$ at almost all time instants, because there is only one unique particle in the set. %in table~\ref{tab:tracking_state_performance}, demonstrating an improvement in accuracy from the smoother.

@@

\includegraphics[width=0.9\columnwidth]{tracking_mtnees.pdf}
\caption{Mean TNEES over 100 scenarios using filter, filter-smoother, and smoother. Optimum value: 0.73}
\label{fig:tracking_mtnees}

@@

Finally, in Fig.~\ref{fig:varying_num_pts} the variation in accuracy and efficiency of the smoothing particle approximation is demonstrated as the number of smoothing particles varies. As expected, the rate of improvement diminishes and becomes negligible as the number of smoothing particles exceeds the number of filtering particles.

@@

\begin{figure}

@@

\subfloat[]{\includegraphics[width=0.45\columnwidth]{MTNEES_vs_numpts.pdf}}
\caption{Average position error (a) and Mean TNEES (b) over 100 scenarios with varying number of smoother particles and 50 filter particles.}
